# (PART\*) Linear models in `R` {-}

# Learning objectives

**Summary :** In this workshop, you will learn the structure of a linear model and its different variants such as simple regression, analysis of variance (ANOVA), analysis of covariance (ANCOVA) and multiple regression. This workshop focuses on implementing linear models in R with `lm()` and `anova()`, identifying models that do not meet the conditions for application and solving problems.

# Preparing for the workshop

All workshop materials are found at [github.com/QCBSRworkshops/workshop04](https://github.com/QCBSRworkshops/workshop04). This includes an [R script](https://r.qcbs.ca/workshop04/book-en/workshop04-script-en.R) which contains all code chunks shown in this book.

For this workshop, we will be working with the following datasets:

* [birdsdiet](https://github.com/QCBSRworkshops/workshop04/edit/main/book-en/data/birdsdiet.csv)
* [dickcissel](https://github.com/QCBSRworkshops/workshop04/edit/main/book-en/data/dickcissel.csv)

> To download this data, do right click + save on the page that opens.

You should also make sure you have downloaded, installed, and loaded these packages:

* [dplyr](https://cran.r-project.org/package=dplyr) (to work with data)
* [vegan](https://cran.r-project.org/package=vegan) (for population data analyses)
* [e1071](https://cran.r-project.org/package=e1071) (to test models)
* [MASS](https://cran.r-project.org/package=MASS) (for linear discriminant analysis)
* [car](https://cran.r-project.org/package=car) (for the analysis of variance)
* [effect](https://cran.r-project.org/package=effect) (to visualize models)

To install them from CRAN, run:

```{r, echo = TRUE, eval = TRUE, message=FALSE, warning=FALSE, results = 'hide'}
# Install the required packages
install.packages("dplyr")
install.packages("vegan")
install.packages("e1071")
install.packages("MASS")
install.packages("car")
install.packages("effects")
```

To load these packages, run:

```{r, echo = TRUE, eval = TRUE, message=FALSE, warning=FALSE}
# Load the required packages
library(dplyr)
library(vegan)
library(e1071)
library(MASS)
library(car)
library(effects)
```

