# Analysis of covariance (ANCOVA)
```{r, echo = F, eval = T, message=FALSE}
# data and all requirements
library(car)
library(effects)
bird <- read.csv('data/birdsdiet.csv')

bird$logMass <- log(bird$Mass)
bird$logMaxAbund <- log(bird$MaxAbund)
bird$Diet <- as.factor(bird$Diet)
```



Analysis of covariance (ANCOVA) is a linear model that tests the
influence of one categorical explanatory variable (or more) and one
continuous explanatory variable (or more) on a continuous response
variable. Each level of the categorical variable is described by its own
slope and intercept. In addition to testing if the response variable
differs for at least one level of the categorical variable, ANCOVA also
tests whether the response variable might be influenced by its
relationship with the continuous variable (called the covariate in
ANCOVA), and by any differences between group levels in the way that the
continuous variable influences the response (i.e. the interaction). The
ANCOVA hypotheses are thus: that there is no difference in the mean
among levels of the categorical variable; there is no correlation
between the response variable and the continuous explanatory variable;
there is no interaction between the categorical and continuous
explanatory variables.

### 6.1 Assumptions

As with models seen above, to be valid ANCOVA models must meet the
statistical assumptions of linear models that can be verified using
diagnostic plots. In addition, ANCOVA models must have:

1.  The same value range for all covariates
2.  Variables that are *fixed*
3.  No interaction between categorical and continuous variables

**[Note:]{.ul}** A *fixed* variable is one that you are specifically
interested in (i.e. bird mass). In contrast, a random variable is noise
that you want to control for (i.e. site a bird was sampled in). If you
have random variables, see the workshop on [Linear Mixed-effect
Models](https://github.com/QCBSRworkshops/workshop06)!

### 6.2 Types of ANCOVA

You can have any number of factors and/or covariates, but as their
number increases, the interpretation of results gets more complex.

The most frequently used ANCOVAs are those with:

1.  one covariate and one factor
2.  one covariate and two factors
3.  two covariates and one factor

The different possible goals of the ANCOVA are to determine the effects
of:

1.  the categorical and continuous variables on the response variable
2.  the categorical variable(s) on the response variable(s) after
    removing the effect of the continuous variable
3.  the categorical variable(s) on the relationship between the
    continuous variables(s) and the response variable

Importantly, these goals are only met if there is no significant
interaction between the categorical and continuous variables! Examples
of significant interactions between the categorical and continuous
variables (for an ANCOVA with one factor and one covariate) are
illustrated by the second and third panels below:


```{r,echo=FALSE,fig.height=3, fig.width=10}
## functions
f1 <- function(x, a, b) {
  return(x*a+b)
}
# conf for plot
col = rgb(118, 143, 175, maxColorValue = 255)
x <- 1:20
par(mfrow = c(1, 3), mar = c(1, 1, 6.5, 4))
# plot 1
plot(x, f1(x, a=1.1,b=2), ylim = c(0, 60), type = 'l', lwd = 2.5, xaxt = "n", yaxt = "n", xlab = "", ylab = "", bty='l', col = col)
lines(f1(x, a=1.1,b=17), lwd = 2.5, col = col)
lines(f1(x, a=0.6,b=22), lwd = 2.5, col = col)
lines(f1(x, a=1.1,b=40), lwd = 2.5, col = col)
mtext('One level of the factor\n has a different slope', side = 3, line = 2, cex = 1.5)
# plot 2
plot(x, f1(x, a=.5,b=2), ylim = c(0, 60), type = 'l', lwd = 2.5, xaxt = "n", yaxt = "n", xlab = "", ylab = "", bty='l', col = col)
lines(f1(x, a=1.1,b=17), lwd = 2.5, col = col)
lines(f1(x, a=1.1,b=22), lwd = 2.5, col = col)
lines(f1(x, a=0.01,b=40), lwd = 2.5, col = col)
mtext('Many levels have\n different slopes', side = 3, line = 2, cex = 1.5)
# plot 3
plot(x, f1(x, a=1.1,b=2), ylim = c(0, 60), type = 'l', lwd = 2.5, xaxt = "n", yaxt = "n", xlab = "", ylab = "", bty='l', col = col)
lines(f1(x, a=1.1,b=17), lwd = 2.5, col = col)
lines(f1(x, a=1.1,b=22), lwd = 2.5, col = col)
lines(f1(x, a=1.1,b=40), lwd = 2.5, col = col)
mtext('No interaction', side = 3, line = 2, cex = 1.5)
```

The same logic follows for ANCOVAs with multiple categorical and/or
continuous variables.

### 6.3 Running an ANCOVA

Running an ANCOVA in R is comparable to running a two-way ANOVA, using
the function `lm`. However, instead of using two categorical variables
(Diet and Aquatic), we now use one categorical and one continuous
variable.

For example, using a build in dataset called CO2, where the response
variable is *uptake*, the continuous variable is *conc* and the factor
is *Treatment*, the ANCOVA is:

```{r, echo = TRUE, eval = TRUE}
ancova.example <- lm(uptake ~ conc*Treatment, data=CO2)
anova(ancova.example)
```

If only your categorical variable is significant, drop your continuous
variable from the model: you will then have an **ANOVA**.\
If only your continuous variable is significant, drop your categorical
variable from the model, you will then have a **simple linear
regression**.\
If your interaction is significant, you might want to test which levels
of your categorical variables ha(s)ve different slopes and to question
whether ANCOVA is the most appropriate model.\
In the CO2 example above, both the continuous and categorical variable
are significant, but the interaction is non-significant. If your replace
*Treatment* with *Type*, however, you will see an example of a
significant interaction.

If you want to compare means across factors, you can use adjusted means,
which uses the equations given by the ANCOVA to estimate the means of
each level of the categorical variable, corrected for the effect of the
categorical variable:

```{r, echo = TRUE, eval = FALSE}
install.packages("effects")
library(effects)
```

```{r, echo = TRUE, eval = TRUE, message=FALSE}
adj.means <- effect('Treatment', ancova.example)
plot(adj.means)

adj.means <- effect('conc*Treatment', ancova.example)
plot(adj.means)
```

------------------------------------------------------------------------

CHALLENGE 4

Run an ANCOVA to test the effect of *Diet*, *Mass*, and their
interaction on *MaxAbund*.\
<details>
  <summary>**Cliquez pour voir la solution au Défi 3!**</summary>

```{r, echo = TRUE, eval = TRUE}
# If you did the section on Contrasts above, you will need to reset the contrast to Treatment for ease of
# comparison using the ''options()'' function
# Otherwise, skip the first line of code below
options(contrasts=c("contr.treatment", "contr.poly"))
ancov1 <- lm(logMaxAbund ~ logMass*Diet, data=bird)
summary(ancov1)
anova(ancov1)
```

In this case, the interaction term is not significant, meaning that the
effect of *logMass* on *logMaxAbund* does not differ between Diet
groups. The interaction term can thus be dropped, and the ANCOVA model
becomes:

```{r, echo = TRUE, eval = FALSE}
ancov2 <- lm(logMaxAbund ~ logMass + Diet, data=bird)
```

Because the R output shows that diet is also non-significant, this term
is dropped, and the final model corresponds to:

```{r, echo = TRUE, eval = FALSE}
lm2 <- lm(logMaxAbund ~ logMass, data=bird)
```

The model results can also be graphically represented using a plot of
the response variable as a function of the continuous explanatory
variable with different point and line colours for the different levels
of the categorical variable.

Plot the ANCOVA intercept and slopes (model ancov1 above) using the
`abline()` and `coef()` functions.

```{r, echo = TRUE, eval = TRUE}
coef(ancov1)


plot(logMaxAbund~logMass, data=bird, col=Diet, pch=19, ylab=expression("log"[10]*"(Maximum Abundance)"),
     xlab=expression("log"[10]*"(Mass)"))
abline(a=coef(ancov1)[1],b=coef(ancov1)[2], col="deepskyblue1")
abline(a=sum(coef(ancov1)[1]+coef(ancov1)[3]),b=sum(coef(ancov1)[2]+coef(ancov1)[7]),col="green2", lwd=2)
abline(a=sum(coef(ancov1)[1]+coef(ancov1)[4]),b=sum(coef(ancov1)[2]+coef(ancov1)[8]),col="orange1", lwd=2)
abline(a=sum(coef(ancov1)[1]+coef(ancov1)[5]),b=sum(coef(ancov1)[2]+coef(ancov1)[9]),col="lightsteelblue1",
       lwd=2)
abline(a=sum(coef(ancov1)[1]+coef(ancov1)[6]),b=sum(coef(ancov1)[2]+coef(ancov1)[10]),col="darkcyan", lwd=2)
```

</details>